Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-15 paper/paper.tex
WIP on paper

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of \citet{simpson_simple_2011} provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in their model is
given instantaneous knowledge of the vector that would move it toward its pre
specified destination. In addition to the globally supervised chemoaffinity
term, \citet{simpson_simple_2011} introduced a competitive interaction for
space between growth cone agents and a receptor-ligand axon-axon interaction
in order to account for the full set of experimental manipulations. Here, we
propose the replacement of the chemoaffinity term with a gradient following
model consisting of axonal growth cone agents which carry receptor molecule
expression determined by their soma's location of origin on the retina. Growth
cones move on the simulated tectum guided by two pairs of opposing, orthogonal
signalling molecules representing the Ephrin ligands. We show that with only
the chemoaffinity term and a receptor-ligand based axon-axon interaction term
(meaning that all growth cone interactions are by receptor-ligand signalling),
a full range of experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
and essential requirement for axons to find their
way \citep{gosse_retinotopic_2008} is also accounted for by the model, due to
the opposing influences of signalling gradient pairs. Finally, we demonstrate
that, assuming exponentially varying receptor expression in the retina, ligand
expression should either be exponential if the receptor-ligand signal induces
repulsion (i.e. gradient descent) or logarithmic if the signal induces
attraction (gradient ascent). Thus, we find that a model analogous to the one
we presented in \citet{james_modelling_2020} that accounts for murine barrel
patterning is also a candidate mechanism for the arrangement of the more
continuous retinotectal system.}

@@

The retinotectal projection has proved to be a deep mine of information for
the study of how the cells of the central nervous system are accurately
connected together into functional networks. This projection connects the
light-gathering cells in the retina to movement-related cells in the optic
tectum (known as the superior colliculus in mammals). Light sources
originating close to each other in the environment tend to activate retinal
cells situated close together in the eye so that an image of the environment
is formed on the retinal surface. It has been discovered that the topography
of the retina is preserved within brain regions that process this information
such that cells which are adjacent within the retina primarily excite cells
adjacent in the tectum. This indicates that during development there must
exist a mechanism which ensures the correct arrangement of the axons which
leave the retina and connect to cells in the tectum.

@@

In our model, space-based competition was retained (though sometimes omitted)
and chemoaffinity and axon-axon interactions were remodelled so that all
interactions occurred via signal transmission between receptors and ligands,
both of which are expressed in graded patterns on both retina and
tectum. Signals induce attractive or repulsive interactions; for chemoaffinity
the interaction is to ascend or descend the gradient of ligand expression on
the tectum; for axon-axon interactions, the strength of short-range repulsion
(or attraction) is governed by the receptor-ligand expression on pairs of
axons.

@@

Each retinal ganglion cell projects $n$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

@@

Figure~\ref{f:ex} shows colour maps of receptor and ligand expression, in this
case tectal ligand expression maps are linear. Note that the expression is
discretized, and in simulations, the gradient is computed numerically from the
discretized expression. This makes it straightforward to perform tectal and
retinal `graft' manipulations. Coordinates within the simulated retina and
tectum are normalised, with 1 corresponding to 2 mm on the tectum and on the
retina for mouse \citep{reber_relative_2004}. This means that growth cones of
approximate radius 5 $\mu$m = 0.005 mm \citep{goodhill_molecular_2016} have a
radius $r$ = 0.0025 in the arbitrary units of the simulation. For a realistic
simulation based on mouse, the approximate RGC density in the retina is around
44000 cells per retina \citep{jeon_major_1998}. This sets the side length of
the retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, axon-axon competition, $\mathbf{C}_b$, an axon-axon
receptor-receptor interaction, $\mathbf{I}_b$, and axon-axon receptor-ligand
interaction, $\mathbf{J}_b$, and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{C}_b + m_3 \mathbf{I}_b +
m_4 \mathbf{J}_b + m_5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$--$m_5$ are scalar parameters.

@@

@@

Competition, in the context of this model, is a repulsive effect which acts
between any two branches, whether they originate from different retinal
ganglion cells or from the same cell. It assumes that there is some kind of
self-other signal (\textbf{be sure to discuss possible origins}) which acts to
discourage two cells from approaching each other.
%
\begin{equation}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k \hat{\mathbf{x}}_{kb} W
\end{equation}
where the distance based weight $W$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$.

@@

We model interactions between axon growth cones due to signalling between
receptors of the same type. For these interactions, repulsion causes a
movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
branch $k$ to branch $b$ causing them to move further apart; attraction causes
the opposite movement. The interaction, $\mathbf{I}_b$, acting on branch $b$,
is given by
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k \sum_i \hat{\mathbf{x}}_{kb}\,W_I \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} < 1.1
\end{equation}
%
the weight here is defined as a function of $r_I$:
%
\begin{equation}
W_I(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_I}   & d_{kb} \leq 2r_I \\
     0 & d_{kb} > 2r_I
     \end{cases}
\end{equation}
$r_{i,k}$ is the expression of receptor $i$ on branch $k$.

@@

We implement a border effect based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary

@@

@@

\subsection*{Observations/Notes}

@@

>>>>

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of \citet{simpson_simple_2011} provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in that model is
given instantaneous knowledge of the vector that would move it toward its
pre-specified destination. In addition to the globally supervised
chemoaffinity term, \citet{simpson_simple_2011} introduced a competitive
interaction for space between growth cone agents and a receptor-receptor
axon-axon interaction in order to account for the full set of experimental
manipulations. Here, we propose the replacement of the chemoaffinity term with
a gradient following model consisting of axonal growth cone agents which carry
receptor molecule expression determined by their soma's location of origin on
the retina. Growth cones move on the simulated tectum guided by two pairs of
opposing, orthogonal signalling molecules representing the Ephrin ligands. We
show that with only the receptor-ligand mediated chemoaffinity term and a
receptor-receptor based axon-axon interaction term (meaning that all growth
cone interactions are communicated by receptor signalling), a full range of
experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
an essential requirement for axons to find their
way \citep{gosse_retinotopic_2008} is also accounted for by the model, due to
the opposing influences of signalling gradient pairs. Finally, we demonstrate
that, assuming exponentially varying receptor expression in the retina, tectal
ligand expression should either be exponential if the receptor-ligand signal
induces repulsion (i.e. gradient descent) or logarithmic if the signal induces
attraction (gradient ascent). Thus, we find that a model analogous to the one
we presented in \citet{james_modelling_2020} that accounts for murine barrel
patterning is also a candidate mechanism for the arrangement of the more
continuous retinotectal system.}

@@

The retinotectal projection has been an important model system in the study of
how the cells of the central nervous system are accurately connected together
into functional networks. This projection connects the light-gathering cells
in the retina to movement-related cells in the optic tectum (known as the
superior colliculus in mammals). Light sources originating close to each other
in the environment tend to activate retinal ganglion cells (either
directly \cite{iRGC_citation} or indirectly via rods and cones) situated close
together in the eye so that an image of the environment is formed on the
retinal surface. It has been discovered that the topography of the retina is
preserved within brain regions that process this information such that cells
which are adjacent within the retina primarily excite cells adjacent in the
tectum. This indicates that during development there must exist a mechanism
which ensures the correct arrangement of the axons which leave the retina and
connect to cells in the tectum.

@@

We sought to replace the globally-supervised mechanisms by local receptor and
ligand based interactions. First, chemoaffinity was remodelled so that all
interactions occurred via signal transmission between the receptors expressed
on the growth cones of retinal ganglion cell axons and the ligands expressed
across the tectum. Signals induced either an attractive or repulsive
interaction to cause growth cones to ascend or descend the gradient of ligand
expression on the tectum.

We then explored the behaviour of models incorporating this chemoaffinity
mechanism alone and in conjunction with several competition mechanisms. We
found parameters for the best models of each type (chemo-only [G],
chemo-competition [GC], chemo-receptor-receptor [GI] and chemo-receptor-ligand
[GJ]) by an optimisation approach. We then compared each model's ability to
reproduce the surgical and genetic manipulation results in a manner similar to
that taken by \citet{simpson_simple_2011} to find the simplest model that
would fit the data.

@@

Each retinal ganglion cell projects $n$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and expressed at levels determined by the cell's
location on the retina. The expression levels of the receptors and ligands
vary with respect to the cell's position on the retinal surface. Each receptor
or ligand varies only with respect to one dimension.  Receptor expression
gradients are arranged in orthogonal pairs with the gradient of $r_0$ being
orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, is
orthogonal to $r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$
opposing $r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to
$l_0$ and opposing $r_1$.

@@

Figure~\ref{f:ex} shows colour maps of receptor and ligand expression, in this
case tectal ligand expression maps are linear. Note that the expression is
discretized, and in simulations, the gradient is computed numerically from the
discretized expression. This makes it straightforward to perform tectal and
retinal `graft' manipulations. Coordinates within the simulated retina and
tectum are normalised, with 1 corresponding to 2 mm on the tectum and on the
retina for mouse \citep{reber_relative_2004}. This means that growth cones of
approximate radius 5 $\mu$m = 0.005 mm \citep{goodhill_can_2016} have a
radius $r$ = 0.0025 in the arbitrary units of the simulation. For a realistic
simulation based on mouse, the approximate RGC density in the retina is around
44000 cells per retina \citep{jeon_major_1998}. This sets the side length of
the retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, axon-axon competition, $\mathbf{C}_b$, an axon-axon
receptor-receptor interaction, $\mathbf{I}_b$, and axon-axon receptor-ligand
interaction, $\mathbf{J}_b$, and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{C}_b + m_3 \mathbf{I}_b +
m_4 \mathbf{J}_b + m_5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$--$m_5$ are scalar parameters. We typically
considered models for which some of $m_2$ to $m_4$ were fixed to zero. A model
for which $m_3$ and $m_4$ were zero was named a `$\mathbf{GC}$' model as it
combined the chemoaffinity effect and space-based competition; a model for
which $m_2$ was zero was a `$\mathbf{GIJ}$' model and so on.

@@

We investigated three forms of competition that could be combined with the
chemoaffinity effect. We investigated space-based competition
(`$\mathbf{C}$'), as described in \citet{simpson_simple_2011}, a
receptor-receptor signalling mechanism (`$\mathbf{I}$'), similar to that
described in \citet{simpson_simple_2011}, and a receptor-ligand mass-action
signalling model (`$\mathbf{J}$'). Our aim was to determine whether either of
the receptor/ligand based signalling mechanisms could replace the space-based
competition.

@@

Space based competition is a repulsive effect which acts between any two
branches, whether they originate from different retinal ganglion cells or from
the same cell. It assumes that there is some kind of self-other signal which
acts to discourage two cells from approaching each other.
%
\begin{equation}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k \hat{\mathbf{x}}_{kb} W
\end{equation}
where the distance based weight $W$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$.

@@

We modelled interactions between axon growth cones due to signalling between
receptors of the same type. For these interactions, repulsion caused a
movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
branch $k$ to branch $b$ causing them to move further apart; attraction caused
the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k \sum_i \hat{\mathbf{x}}_{kb}\,W_I \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} < 1.1
\end{equation}
%
the weight here is defined as a function of $r_I$:
%
\begin{equation}
W_I(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_I}   & d_{kb} \leq 2r_I \\
     0 & d_{kb} > 2r_I
     \end{cases}
\end{equation}
$r_{i,k}$ is the expression of receptor $i$ on branch $k$.

@@

We implemented a border effect based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary

@@

\subsection*{Simulation}

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that \textbf{i)} All
receptor-ligand signalled interactions are repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecexp}.

In all simulations, the border effect, $B_b$, had the same magnitude with
$m_5=0.5$. The number of branches per axon was 4 unless otherwise stated.

@@

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300 steps.

% ./build/sim/agent/agent1 configs/a1/m_ee_G.json  configs/a1/e_wt.json 300
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

@@

\subsection*{Results Observations/Notes}
